---
title: "awesomlab4: A linear regression package"
author: "Martin Svensson and Laura Julià Melis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{awesomelab4: A linear regression package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## 1. Introduction.
The **awesomelab4** package contains an object with class linreg, created with the RC object oriented system. This object is helpful to handle linear regression models as it provides multiple methods such as `resid()` and `coef()` to obtain the vector of residuals and the regressions coefficients of a given regression model, respectively.

## 2. Statistics calculations. 
By default, tha class object calculates various statistics and stores them. This section aims to briefly explain these estimations. Let $\boldsymbol{X}$ be the matrix of independent variables and $\boldsymbol{y}$ the vector with the dependent variable values, then `linreg()` will estimate the following:

**i. Regressions coefficients:**
$$\hat{\beta}=(\boldsymbol{X}^{\boldsymbol{T}}\boldsymbol{X})^{-1} \boldsymbol{X}^{\boldsymbol{T}}\boldsymbol{y}$$
**ii. The fitted values:**
$$\hat{\boldsymbol{y}}= \boldsymbol{X}\hat{\beta}$$
**iii. The residuals:**
$$\boldsymbol{\hat{e}}= \boldsymbol{y}-\boldsymbol{\hat{y}}= \boldsymbol{y}- \boldsymbol{X}\hat{\beta}$$
**iv. The degrees of freedom:**
$$df=n-p$$
where $n$ is the number of observations and $p$ is the number of parameters in the model.

**v. The residual variance:**
$$\hat{\sigma}^2=\frac{\boldsymbol{e}^\boldsymbol{T}\boldsymbol{e}}{df}$$
**vi. The variance of the regression coefficients:**
$$\widehat{\boldsymbol{Var}}(\hat{\beta})=\hat{\sigma}^2(\boldsymbol{X}^{\boldsymbol{T}}\boldsymbol{X})^{-1}$$
**vii. The t-values for each coefficient:**
$$t_{\beta}=\frac{\hat{\beta}}{\sqrt{Var(\hat{\beta})}}$$

## 3. linreg( ) description.
The class object `linreg()` stores the statistics seen in the section above as *fields* and contains multiple functions (*methods*) to obtain outputs related to linear regression models. The name and a brief description of each one is listed below. 

**1. Fields.**
 
  - `formula:` an object of class "formula".
  - `data:` a data frame.
  - `regression_coefficients:` the regression coefficients of the model calculated using the QR decomposition.
  - `fitted_values:` the fitted values.
  - `residuals:` the residuals. It is the difference between the real values and the fitted values.
  - `degrees_of_freedom:` the degrees of freedom: the number of observations minus the number of parameters in the model.
  - `residual_variance:` the residual variance. 
  - `variance_of_coefficients:` the variance of the regression coefficients.
  - `std_error:` the standard error of the regression coefficients. It is the square root of the variance.
  - `t_values:` the t-values for each coefficient.
  - `p_values:` the p-values for each coefficient.

**2. Methods.**
 
  - `initialize()` calculates all the statistics described in section **2** from formula and data.
  - `print()` prints out the coefficients and coefficient names.
  - `plot()` plots two different plots using ggplot2: "Residuals vs Fitted" and "Scale-Location".
  - `resid()` returns the vector of residuals.
  - `pred()` returns the predicted values.
  - `coef()` returns the coefficients as a named vector.

## 4. Code.
```{r message=FALSE, warning=FALSE}
linreg <- setRefClass ("linreg",
  fields = c (
    formula = "formula",
    data = "ANY",
    regression_coefficients = "ANY", 
    fitted_values = "ANY",
    residuals = "ANY",
    std_residuals = "ANY",
    degrees_of_freedom = "ANY",
    residual_variance = "ANY",
    variance_of_coefficients = "ANY",
    std_error = "ANY",
    t_values = "ANY",
    p_values = "ANY",
    arguments = "ANY"),
  methods = c (
    initialize = function(formula, data){
      "Initialize function calculates all values needed from formula and data."
      ## 0. Check that the class of the formula argument is correct:
      stopifnot(class(formula) == "formula")

      # Extract arguments
      arg_string <- paste(as.list(sys.calls()), collapse='')
      
      # Extract function call
      function_pattern <- "linreg\\$new\\(.*?\\)"
      function_call <- regmatches(arg_string, regexpr(function_pattern, arg_string, perl = TRUE))
      
      # From function call, extract data argument
      data_pattern <- "(\\w+)(?=\\))"
      data_argument <- regmatches(function_call, regexpr(data_pattern, function_call, perl = TRUE))
      
      # Save arguments
      arguments <<- c(formula, data_argument)
      
      ## 1. Initialization
      x <- model.matrix(formula, data) # X matrix (independent variables)
      name_dep <- all.vars(formula)[1] # Dependent variable/s name/s
      y <- data[, name_dep] # y (dependent variable/s)
      
      ## 2. Calculation of Q and R
      qr_x <- qr(x)
      Q <- qr.Q(qr_x) # orthogonal matrix
      R <- qr.R(qr_x) # triangular matrix
      
      ## 3. Estimations (Computations using ordinary least squares).
      # Regression coefficients:
      regression_coefficients <<- as.vector(backsolve(R, (t(Q) %*% y)))
      
      # The fitted values:
      fitted_values <<- x %*% regression_coefficients
      
      # The residuals:
      residuals <<- y - (x %*% regression_coefficients)
      std_residuals <<- residuals/(sqrt((1/(nrow(data)-1))*sum(residuals^2)))
      
      # The degrees of freedom:
      degrees_of_freedom <<- nrow(data) - ncol(x) # number of observations - number of parameters in the model
      
      # The residual variance:
      residual_variance <<- as.numeric((t(residuals)%*%residuals)/degrees_of_freedom)
      
      # The variance of the regression coefficients:
      variance_of_coefficients <<- diag(solve(t(R)%*%R) * residual_variance) # Var(beta_hat)= (R^T * R)^(-1) * sigma_hat^2
      std_error <<- t(sqrt(variance_of_coefficients))    # Std. Errors
      
      # The t-values for each coefficient:
      t_values <<- regression_coefficients / std_error
      
      # The p-values for each coefficient:
      #p_values <<- 2*abs(pt(t_values, degrees_of_freedom, log.p = T))
      p_values <<- pt(-abs(t_values), degrees_of_freedom)
      return(NULL)
    },
    show = function() {
      "Modifies the print() function for class."
      base::print("Use ...$print()")
    },
    print = function() {
      "Same as show, but allows for linreg$print(). Will break print(linreg) that show allows, bad test case!"
      coeff <- matrix(c(names(std_error), regression_coefficients), ncol=3)
      colnames(coeff) <- colnames(t_values)
      
      model <- list()
      model$Call <- paste("linreg(formula = ", arguments[1], ", data = ", arguments[2], ")", sep="")
      model$Coefficients <- coeff
      base::print(model)
    },
    plot = function(){
      liu_theme <- theme_bw() + theme(
        plot.title = element_text(color = "#00b9e7"),
        panel.background = element_rect(fill = "#c7e6e3", colour = "#17c7d2",
                                        size = 2, linetype = "solid"),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = "white"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                        colour = "white"),
        #scale_color_manual(values=c('#ff6442','#8781d3', '#fcf05f', '#687f91')),
        legend.title = element_text(face = "bold"),
        legend.position = "bottom",
        axis.text.x = element_text(color="#687f91"),
        axis.text.y = element_text(color="#687f91")
        )
      
      df <- data.frame(residuals, std_residuals, fitted_values)
      plot1 <- ggplot(df, aes(y=residuals, x=fitted_values)) + geom_point() + geom_line(aes(y = mean(residuals), x= fitted_values, colour="red")) + ggtitle("Residuals vs Fitted") + xlab("Fitted values") + ylab("Residuals") + liu_theme
      plot2 <- ggplot(df, aes(y=std_residuals, x=fitted_values)) + geom_point() + geom_line(aes(y = mean(std_residuals), x= fitted_values, colour="red")) + ggtitle("Scale-Location") + xlab("Fitted values") + ylab(expression(sqrt("|Standardized residuals|"))) + liu_theme
      plot_grid(plot1, plot2, labels = "AUTO")
      },
    resid = function(){
      "Returns the residuals."
      return(residuals)
    },
    pred = function(){
      "Returns the predictions."
      return(fitted_values)
    },
    coef = function(){
      "Returns the coefficients."
      return(regression_coefficients)
    },
    summary = function() {
      "Prints a summary for the calculated data."
      .categorise_significance <- function(x){
        if(x < 0.001){
          return("***")
        }else if(x < 0.01){
          return("**")
        }else if(x < 0.05){
          return("*")
        }else if(x < 0.1){
          return(".")
        }else{
          return(" ")
        }
      }
      significance <- lapply(as.vector(p_values), .categorise_significance)
      coef <- cbind(regression_coefficients, std_error[1,], t_values[1,], p_values[1,], significance)
      colnames(coef) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)", "Significance")
      base::print(coef)
      
      base::print("Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1")
      rse <- sqrt(sum(residuals**2)/degrees_of_freedom)
      base::print(paste("Residual standard error: ", rse, " on ", degrees_of_freedom, " degrees of freedom", sep=""))
    }
  )
)
```

## 5. Example.
This section shows how to use `linreg()` on a well known dataset such as iris.
```{r include=FALSE}
library(ggplot2)
library(cowplot)
```

### Initialization.
```{r}
# Loading the dataset.
data(iris) 

# Setting "formula" and "data" arguments
linreg_mod <- linreg$new(Petal.Length~Sepal.Width+Sepal.Length, data=iris)
```

### Methods.
**print()**
```{r}
linreg_mod$print()
```

**plot()**
```{r}
linreg_mod$plot()
```

**resid()**
```{r}
head(linreg_mod$resid())
```

**pred()**
```{r}
head(linreg_mod$pred())
```

**coef()**
```{r}
linreg_mod$coef()
```

## References.
* Reference Classes: <http://adv-r.had.co.nz/R5.html>
* QR decomposition: <https://en.wikipedia.org/wiki/QR_decomposition>
* Vignettes: <http://r-pkgs.had.co.nz/vignettes.html>