---
title: "awesomlab4: A linear regression package"
author: "Martin Svensson and Laura Julià Melis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{awesomelab4: A linear regression package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## 1. Introduction.
The **awesomelab4** package contains an object with class linreg, created with the RC object oriented system. This object is helpful to handle linear regression models as it provides multiple methods such as `resid()` and `coef()` to obtain the vector of residuals and the regressions coefficients of a given regression model, respectively.

## 2. Statistics calculations. 
By default, tha class object calculates various statistics and stores them. This section aims to briefly explain these estimations. Let $\boldsymbol{X}$ be the matrix of independent variables and $\boldsymbol{y}$ the vector with the dependent variable values, then `linreg()` will estimate the following:

**i. Regressions coefficients:**
$$\hat{\beta}=(\boldsymbol{X}^{\boldsymbol{T}}\boldsymbol{X})^{-1} \boldsymbol{X}^{\boldsymbol{T}}\boldsymbol{y}$$
**ii. The fitted values:**
$$\hat{\boldsymbol{y}}= \boldsymbol{X}\hat{\beta}$$
**iii. The residuals:**
$$\boldsymbol{\hat{e}}= \boldsymbol{y}-\boldsymbol{\hat{y}}= \boldsymbol{y}- \boldsymbol{X}\hat{\beta}$$
**iv. The degrees of freedom:**
$$df=n-p$$
where $n$ is the number of observations and $p$ is the number of parameters in the model.

**v. The residual variance:**
$$\hat{\sigma}^2=\frac{\boldsymbol{e}^\boldsymbol{T}\boldsymbol{e}}{df}$$
**vi. The variance of the regression coefficients:**
$$\widehat{\boldsymbol{Var}}(\hat{\beta})=\hat{\sigma}^2(\boldsymbol{X}^{\boldsymbol{T}}\boldsymbol{X})^{-1}$$
**vii. The t-values for each coefficient:**
$$t_{\beta}=\frac{\hat{\beta}}{\sqrt{Var(\hat{\beta})}}$$

## 3. linreg( ) description.
The class object `linreg()` stores the statistics seen in the section above as *fields* and contains multiple functions (*methods*) to obtain outputs related to linear regression models. The name and a brief description of each one is listed below. 

**1. Fields.**
 
  - `formula:` an object of class "formula".
  - `data:` a data frame.
  - `regression_coefficients:` the regression coefficients of the model calculated using the QR decomposition.
  - `fitted_values:` the fitted values.
  - `residuals:` the residuals. It is the difference between the real values and the fitted values.
  - `degrees_of_freedom:` the degrees of freedom: the number of observations minus the number of parameters in the model.
  - `residual_variance:` the residual variance. 
  - `variance_of_coefficients:` the variance of the regression coefficients.
  - `std_error:` the standard error of the regression coefficients. It is the square root of the variance.
  - `t_values:` the t-values for each coefficient.
  - `p_values:` the p-values for each coefficient.

**2. Methods.**
 
  - `initialize()` calculates all the statistics described in section **2** from formula and data.
  - `print()` prints out the coefficients and coefficient names.
  - `plot()` plots two different plots using ggplot2: "Residuals vs Fitted" and "Scale-Location".
  - `resid()` returns the vector of residuals.
  - `pred()` returns the predicted values.
  - `coef()` returns the coefficients as a named vector.

## 4. Example.
This section shows how to use `linreg()` on a well known dataset such as iris.
```{r include=FALSE}
library(ggplot2)
library(cowplot)
library(awesomelab4)
```

### Initialization.
```{r}
# Loading the dataset.
data(iris) 

# Setting "formula" and "data" arguments
linreg_mod <- linreg$new(Petal.Length~Sepal.Width+Sepal.Length, data=iris)
```

### Methods.
**print()**
```{r}
linreg_mod$print()
```

**plot()**
```{r plot1, fig.height=4, fig.width=7, fig.align="center"}
plot1 <- linreg_mod$plot()
plot1
```

**resid()**
```{r}
head(linreg_mod$resid())
```

**pred()**
```{r}
head(linreg_mod$pred())
```

**coef()**
```{r}
linreg_mod$coef()
```

## References.
* Reference Classes: <http://adv-r.had.co.nz/R5.html>
* QR decomposition: <https://en.wikipedia.org/wiki/QR_decomposition>
* Vignettes: <http://r-pkgs.had.co.nz/vignettes.html>